cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated sphere data generation

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane:

$$
X_1 = \cos(\theta), \quad X_2 = \sin(\theta)
$$

For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$.

Each additional dimension is computed as:

$$
X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p
$$

This structure retains a one-dimensional manifold (the circle) but allows it to be embedded non-linearly and decorrelatively in higher dimensions. The decreasing scale ensures that higher dimensions add less variation, preserving the dominance of the circular shape while adding complexity.

@@

Together, these define a periodic, non-trivial, closed curve in $3\text{-}D$ with internal folds that produce a more complex geometry than a standard circle or ellipse.

For dimensions $X_4$ through $X_p$, additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as $\text{scale}_j = \sqrt{(0.5)^{j-3}}$ for $j$ ranging from 4 to $p$, which means that the amplitude decreases as the dimension increases. Each dimension $X_j$ is then calculated using the formula $X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)$, where the phase shift $\phi_j$ is given by $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$.

This introduces decorrelated, low-variance sinusoidal variations across dimensions, retaining the cyclical nature while embedding the data in higher-dimensional space with a coherent but subtle structure.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle.

The corresponding Cartesian coordinates are calculated as;

$$
X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta),
$$
$$
X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta),
$$
$$
X_3 = r \cdot u,
$$

which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within).

For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise:

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

This extends the $3\text{-}D$ spherical manifold into $p$-dimensional space with orthogonal Gaussian noise, preserving the spherical structure while allowing for testing in higher-dimensional settings.

@@

The base structure consists of a $3\text{-}D$ spherical surface created using a regular grid in spherical coordinates, where $\theta \in [0, 2\pi]$ represents the azimuthal angle (longitude) and $\phi \in [0, \pi]$ denotes the polar angle (co-latitude).

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation:

$$
X_1 = \sin(\phi) \cdot \cos(\theta),
$$
$$
X_2 = \sin(\phi) \cdot \sin(\theta),
$$
$$
X_3 = \cos(\phi).
$$

This forms a structured grid of points on the unit sphere, ideal for studying how regular geometric manifolds are transformed under nonlinear dimensionality reduction.

For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions:

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

These dimensions represent low-variance orthogonal noise, preserving the spherical shape while testing robustness in high-dimensional settings.

@@

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + k \cdot n_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location.

A large uniform sphere centered at the origin is created by sampling $n_1$ points uniformly on the surface of a $p$-dimensional sphere with a radius of $r_1$. The sampling is executed using the function `gen_unifsphere(n_1, p, r_1)`, which generates the desired points in the specified dimensional space.

@@

Points on spheres are generated using the standard hyperspherical method, which involves sampling $u \sim U(-1, 1)$ to determine the cosine of the polar angle, and sampling $\theta \sim U(0, 2\pi)$ to determine the azimuthal angle (for 3D).

Each observation is classified by cluster, with labels such as "big" for the large central sphere and "small_1" to "small_k" for the smaller spheres.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere.

The coordinates are transformed into $4\text{-}D$ Cartesian space:

$$
X_1 = \sin(\theta_1) \cdot \cos(\theta_2),
$$
$$
X_2 = \sin(\theta_1) \cdot \sin(\theta_2),
$$
$$
X_3 = \cos(\theta_1) \cdot \cos(\theta_3),
$$
$$
X_4 = \cos(\theta_1) \cdot \sin(\theta_3).
$$

This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere.

For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise:

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p.
$$

These dimensions maintain the spherical structure while simulating embedding into higher-dimensional space with small orthogonal perturbations.

>>>>

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$. Each additional dimension is computed as: $X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

Together, these define a periodic, non-trivial, closed curve in $3\text{-}D$ with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions $X_4$ through $X_p$, additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as $\text{scale}_j = \sqrt{(0.5)^{j-3}}$ for $j$ ranging from $4$ to $p$, which means that the amplitude decreases as the dimension increases. Each dimension $X_j$ is then calculated using the formula $X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)$, where the phase shift $\phi_j$ is given by $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta)$, $X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta)$, $X_3 = r \cdot u$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

The base structure consists of a $3\text{-}D$ spherical surface created using a regular grid in spherical coordinates, where $\theta \in [0, 2\pi]$ represents the azimuthal angle and $\phi \in [0, \pi]$ denotes the polar angle.

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: $X_1 = \sin(\phi) \cdot \cos(\theta)$, $X_2 = \sin(\phi) \cdot \sin(\theta)$, $X_3 = \cos(\phi)$. For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

The `gen_clusteredspheres(n, k, p, r, loc)` function generates a synthetic dataset of $n_1 + kn_2$ observations in $p$-dimensional space, consisting of one large sphere of radius $r_1$ and $k$ smaller spheres of radius $r_2$, each centered at a different random location.

A large uniform sphere centered at the origin is created by sampling $n_1$ points uniformly on the surface of a $p\text{-}D$ sphere with a radius of $r_1$. The sampling is executed using the function `gen_unifsphere(n_1, p, r_1)`, which generates the desired points in the specified dimensional space.

@@

Points on spheres are generated using the standard hyperspherical method, which involves sampling $u \sim U(-1, 1)$ to determine the cosine of the polar angle, and sampling $\theta \sim U(0, 2\pi)$ to determine the azimuthal angle (for $3\text{-}D$). Each observation is classified by cluster, with labels such as "big" for the large central sphere and "small_1" to "small_k" for the smaller spheres.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1) \cdot \cos(\theta_2)$, $X_2 = \sin(\theta_1) \cdot \sin(\theta_2)$, $X_3 = \cos(\theta_1) \cdot \cos(\theta_3)$, $X_4 = \cos(\theta_1) \cdot \sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p$.

### 2025-06-24 paper-cardinalR.tex
Updated sphere data generation

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane:

\[
X_1 = \cos(\theta), \quad X_2 = \sin(\theta)
\]

For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\).

Each additional dimension is computed as:

\[
X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p
\]

This structure retains a one-dimensional manifold (the circle) but allows it to be embedded non-linearly and decorrelatively in higher dimensions. The decreasing scale ensures that higher dimensions add less variation, preserving the dominance of the circular shape while adding complexity.

@@

Together, these define a periodic, non-trivial, closed curve in \(3\text{-}D\) with internal folds that produce a more complex geometry than a standard circle or ellipse.

For dimensions \(X_4\) through \(X_p\), additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as \(\text{scale}_j = \sqrt{(0.5)^{j-3}}\) for \(j\) ranging from 4 to \(p\), which means that the amplitude decreases as the dimension increases. Each dimension \(X_j\) is then calculated using the formula \(X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)\), where the phase shift \(\phi_j\) is given by \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\).

This introduces decorrelated, low-variance sinusoidal variations across dimensions, retaining the cyclical nature while embedding the data in higher-dimensional space with a coherent but subtle structure.

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle.

The corresponding Cartesian coordinates are calculated as;

\[
X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta),
\]
\[
X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta),
\]
\[
X_3 = r \cdot u,
\]

which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within).

For \(p > 3\), additional dimensions \(X_4\) to \(X_p\) are generated as low-variance Gaussian noise:

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
\]

This extends the \(3\text{-}D\) spherical manifold into \(p\)-dimensional space with orthogonal Gaussian noise, preserving the spherical structure while allowing for testing in higher-dimensional settings.

@@

The base structure consists of a \(3\text{-}D\) spherical surface created using a regular grid in spherical coordinates, where \(\theta \in [0, 2\pi]\) represents the azimuthal angle (longitude) and \(\phi \in [0, \pi]\) denotes the polar angle (co-latitude).

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation:

\[
X_1 = \sin(\phi) \cdot \cos(\theta),
\]
\[
X_2 = \sin(\phi) \cdot \sin(\theta),
\]
\[
X_3 = \cos(\phi).
\]

This forms a structured grid of points on the unit sphere, ideal for studying how regular geometric manifolds are transformed under nonlinear dimensionality reduction.

For \(p > 3\), the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions:

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
\]

These dimensions represent low-variance orthogonal noise, preserving the spherical shape while testing robustness in high-dimensional settings.

@@

The \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates a synthetic dataset of \(n_1 + k \cdot n_2\) observations in \(p\)-dimensional space, consisting of one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location.

A large uniform sphere centered at the origin is created by sampling \(n_1\) points uniformly on the surface of a \(p\)-dimensional sphere with a radius of \(r_1\). The sampling is executed using the function \texttt{gen\_unifsphere(n\_1,\ p,\ r\_1)}, which generates the desired points in the specified dimensional space.

@@

Points on spheres are generated using the standard hyperspherical method, which involves sampling \(u \sim U(-1, 1)\) to determine the cosine of the polar angle, and sampling \(\theta \sim U(0, 2\pi)\) to determine the azimuthal angle (for 3D).

Each observation is classified by cluster, with labels such as ``big'' for the large central sphere and ``small\_1'' to ``small\_k'' for the smaller spheres.

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere.

The coordinates are transformed into \(4\text{-}D\) Cartesian space:

\[
X_1 = \sin(\theta_1) \cdot \cos(\theta_2),
\]
\[
X_2 = \sin(\theta_1) \cdot \sin(\theta_2),
\]
\[
X_3 = \cos(\theta_1) \cdot \cos(\theta_3),
\]
\[
X_4 = \cos(\theta_1) \cdot \sin(\theta_3).
\]

This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere.

For \(p > 4\), additional dimensions \(X_5\) to \(X_p\) are added as low-variance Gaussian noise:

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p.
\]

These dimensions maintain the spherical structure while simulating embedding into higher-dimensional space with small orthogonal perturbations.

>>>>

A latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\). Each additional dimension is computed as: \(X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

Together, these define a periodic, non-trivial, closed curve in \(3\text{-}D\) with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions \(X_4\) through \(X_p\), additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as \(\text{scale}_j = \sqrt{(0.5)^{j-3}}\) for \(j\) ranging from \(4\) to \(p\), which means that the amplitude decreases as the dimension increases. Each dimension \(X_j\) is then calculated using the formula \(X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)\), where the phase shift \(\phi_j\) is given by \(\phi_j = (j - 2) \cdot \frac{\pi}{2p}\).

@@

Each observation lies on the surface of a sphere in \(3\text{-}D\), constructed by generating \(u \sim U(-1, 1)\) which represents the cosine of the polar angle \(\phi\) and \(\theta \sim U(0, 2\pi)\) which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; \(X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta)\), \(X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta)\), \(X_3 = r \cdot u\), which gives points uniformly distributed on the surface of a \(3\text{-}D\) sphere (not within). For \(p > 3\), additional dimensions \(X_4\) to \(X_p\) are generated as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

The base structure consists of a \(3\text{-}D\) spherical surface created using a regular grid in spherical coordinates, where \(\theta \in [0, 2\pi]\) represents the azimuthal angle and \(\phi \in [0, \pi]\) denotes the polar angle.

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: \(X_1 = \sin(\phi) \cdot \cos(\theta)\), \(X_2 = \sin(\phi) \cdot \sin(\theta)\), \(X_3 = \cos(\phi)\). For \(p > 3\), the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

@@

The \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates a synthetic dataset of \(n_1 + kn_2\) observations in \(p\)-dimensional space, consisting of one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location.

A large uniform sphere centered at the origin is created by sampling \(n_1\) points uniformly on the surface of a \(p\text{-}D\) sphere with a radius of \(r_1\). The sampling is executed using the function \texttt{gen\_unifsphere(n\_1,\ p,\ r\_1)}, which generates the desired points in the specified dimensional space.

@@

Points on spheres are generated using the standard hyperspherical method, which involves sampling \(u \sim U(-1, 1)\) to determine the cosine of the polar angle, and sampling \(\theta \sim U(0, 2\pi)\) to determine the azimuthal angle (for \(3\text{-}D\)). Each observation is classified by cluster, with labels such as ``big'' for the large central sphere and ``small\_1'' to ``small\_k'' for the smaller spheres.

@@

Each observation is situated on a restricted \(4\text{-}D\) spherical surface, defined by spherical coordinates. The azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1) \cdot \cos(\theta_2)\), \(X_2 = \sin(\theta_1) \cdot \sin(\theta_2)\), \(X_3 = \cos(\theta_1) \cdot \cos(\theta_3)\), \(X_4 = \cos(\theta_1) \cdot \sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional dimensions \(X_5\) to \(X_p\) are added as low-variance Gaussian noise: \(X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p\).
